An Algorithm for Searching Evolving Graphs in Linear Algebraic Language

### 2016-01-23 paper.tex
Fix fencepost error in root
Reorganize the analysis section

\begin{algorithm}[h]
 \SetKwFor{For}{for}{}{end}
 \SetKwFor{While}{while}{}{end}
 \SetKwFor{If}{if}{}{end}
 \SetKwFunction{nei}{forwardneighbors}
 \SetKwProg{Fn}{function}{}{end}
 \Fn{BFS($G_n, (v_0,t_0)$)}{
  $reached[(v_0,t_0)] = 0$ \\
  $frontier = \{(v_0,t_0)\}$ \\
  $k = 1$ \\
  \While{$frontier \ne \emptyset$}{
   $nextfrontier = \emptyset$ \\
   \For{$(v,t) \in frontier$}{
     \For{$(v',t') \in$\nei{$(v,t)$}}{
          \If{$(v',t') \notin d$}{
              $reached[(v',t')] = k$\\
              $nextfrontier = nextfrontier \cup \{(v',t')\}$ \\
              }
         }
       }
  $frontier = nextfrontier$ \\
  $k = k+1$ \\
  }
  \textbf{return} $reached$
}
\caption{Breadth-first search (BFS) on an evolving graph $G_n$ starting from a
root $(v_1, t_1)$. The return value, $reached$, is a dictionary mapping all
reachable temporal notes from the root to their distances from the root. At the
end of each iteration $k$, the $frontier$ set contains all temporal nodes of
distance $k$ fro the root.
}
\label{alg:bfs}
\end{algorithm}

%Also need to consider the operations for matrix lists.

\subsection{Algebraic Evolving-Graph BFS}
\label{sec:representation}
As before, let $G_n = \langle G^{[1]}, G^{[2]}, \ldots, G^{[n]}\rangle$ be an evolving graph with $n$ time stamps. For each graph $G^{[t]}$, we use $A^{[t]}$ to denote
the corresponding adjacency matrix.
\[
A^{[t]}(i,j) =
\begin{cases}
 1 & \mbox{if $(i,j)$ is an edge of $G^{[t]}$,} \\
 0 & \mbox{otherwise.}
\end{cases}
\]
 Each $A^{[t]}$ can be represented by
a $|V|\times |V|$ sparse matrix,
where $|V|$ is the total number of nodes in $G_n$.
We use a sequence of adjacency matrices $A_n = \langle A^{[1]}, A^{[2]}, \ldots, A^{n]}\rangle$ to represent $G_n$.
For example, the evolving graph in Figure \ref{fig:eg_shortest_path}
can be represented as
\[
\left\langle
  \begin{bmatrix}
    0 & 1 & 0 \\
    0 & 0 & 0 \\
    0 & 0 & 0
  \end{bmatrix},~
 \begin{bmatrix}
   0 & 0 & 1 \\
   0 & 0 & 0 \\
   0 & 0 & 0
 \end{bmatrix},~
 \begin{bmatrix}
  0 & 0 & 0 \\
  0 & 0 & 1 \\
  0 & 0 & 0
 \end{bmatrix}
\right\rangle
\]

@@

% adjacency matrices
Suppose $G_n$ is represented by adjacency lists.
Then the number of adjacency lists is the same as the number
of active nodes in $G_n$. So for large $n$, this representation
can use a large amount of memory.
For the graph $G^{[i]} = (V^{[i]}, E^{[i]})$ at time stamp $i$,
we denote the active node set of $V^{[i]}$ as $\tilde V^{[i]}$,
where we have $\tilde V^{[i]} \subseteq V^{[i]}$.
Step $10$ of  Algorithm \ref{alg:bfs} ensures each active node-time stamp pair in $G_n$ is enqueued in $next$ (and then $frontier$) at most once.
And so \texttt{forward-neighbors} loops through each
active node-time stamp pair at most once.
The operations of enqueuing take $O(1)$ time.
Thus the total operations of enqueuing are
$O(\sum_{i=1}^n|\tilde V^{[i]}|)$, where $|\cdot|$ represents the number of elements of a set.  Let $\sum_i |\tilde V^{[i]}| = |\tilde V|$ and $\sum_i |E^{[i]}| = |\hat E|$.
Then the time of scanning
\texttt{forward-neighbors} is
\[
\sum_{i=1}^n\sum_{v \in \tilde V^{[i]}} \texttt{forward-neighbors}(v) =
O(|\hat E|),
\]
by the definition of forward-neighbors.
Hence the total running time is $O(|\hat E|+ |\tilde V|)$.

\begin{thm}[Correctness of evolving-graph BFS]
Let an evolving graph $G_n$ and a starting node $(s,t_0)$ be given.
The evolving-graph BFS discovers every active node that is reachable from the source $(s,t_0)$ and $dict.(v, t)$ is the length of the shortest temporal path
from $(s,t_0)$ to $(v,t)$.
\end{thm}
\proof
We could map an evolving graph $G_n$ to a static graph $G$ in such a way that
if there is an edge from $(v, t)$ to $(u,t)$ in $G_n$, there is an edge from $v_t$ to $u_t$ in $G$.
For any two nodes $v_s$ and $v_t$ in $G$,  if $s < t$, we add an edge from $v_s$ to $v_t$.
By construction, an active node $(v,t)$ in $G_n$ is in one-to-one correspondence with
a node $v_t$ in $G$.
The forward-neighbors of an active node $(v, t)$ in $G_n$
are in one-to-one correspondence with the forward-neighbors of a node $v_t$ in $G$.
Hence the correctness of the evolving graph BFS follows from that of a static graph.
\endproof

@@

\subsection{Search Tree}
\label{sec:short-temp-path}

We say two nodes $(v_1,t_1)$ and $(v2, t_2)$ are \emph{temporally connected} if
$(v_1, t_1)$ and $(v2,t_2)$ are connected by a temporal path.
Notice this notion is symmetric, reflexive but not transitive. For example, suppose
$t_1 < t_2 < t_3$ and $(v_1, t_1)$ and $(v_3, t_3)$ are temporally connected
by a temporal path from $(v_1,t_1)$ to $(v_3, t_3)$. Also
$(v_3, t_3)$ and $(v_2, t_2)$ are temporally connected by a temporal path from
$(v_2,t_2)$ to $(v_3, t_3)$. But this does not imply $(v_1, t_1)$ and $(v_2, t_2)$ are temporal connected.

Given a root node $(s, t)$, during the execution the evolving-graph BFS builds a search tree, denoted by $T(s,t)$,
where nodes in $T(s,t)$ are temporally connected.
\begin{lemma}
For a node active at different time stamps, say $(s, t_1)$ and $(s,t_2)$ such that $t_1 \leq t_2$,  we have $T(s, t_2) \subseteq T(s, t_1)$, where $T(s,t_1) = T(s,t_2)$ if and only if
$t_1 = t_2$.
\end{lemma}
\proof
Trivially, $t_1 =t_2$ if and only if $(s,t_1) = (s,t_2)$ if and only if
$T(s,t_1) = T(s,t_2)$. Now suppose $t_1 < t_2$.
Let $(v, t)$ be an arbitrary active node that belongs to $T(s, t_2)$. Then there exists a temporal path
(constructed by the evolving graph BFS) from $(s,t_2)$ to $(v,t)$. Since both
$(s,t_1)$ and $(s,t_2)$ are active, $(s,t_2)$ is a forward-neighbor of $(s,t_1)$ by definition, i.e., there exists an edge from $(s,t_1)$ to $(s,t_2)$.  Hence there exists a temporal path
from $(s,t_1)$ to $(v,t)$ and so $(v,t) \in T(s,t_1)$.
\endproof

@@

>>>>

\begin{algorithm}[h]
 \SetKwFor{For}{for}{}{end}
 \SetKwFor{While}{while}{}{end}
 \SetKwFor{If}{if}{}{end}
 \SetKwFunction{nei}{forwardneighbors}
 \SetKwProg{Fn}{function}{}{end}
 \Fn{BFS($G_n, (v_1,t_1)$)}{
  $reached[(v_1,t_1)] = 0$ \\
  $frontier = \{(v_1,t_1)\}$ \\
  $k = 1$ \\
  \While{$frontier \ne \emptyset$}{
   $nextfrontier = \emptyset$ \\
   \For{$(v,t) \in frontier$}{
     \For{$(v',t') \in$\nei{$(v,t)$}}{
          \If{$(v',t') \notin d$}{
              $reached[(v',t')] = k$\\
              $nextfrontier = nextfrontier \cup \{(v',t')\}$ \\
              }
         }
       }
  $frontier = nextfrontier$ \\
  $k = k+1$ \\
  }
  \textbf{return} $reached$
}
\caption{Breadth-first search (BFS) on an evolving graph $G_n$ starting from a
root $(v_1, t_1)$. The return value, $reached$, is a dictionary mapping all
reachable temporal notes from the root to their distances from the root. At the
end of each iteration $k$, the $frontier$ set contains all temporal nodes of
distance $k$ from the root.
}
\label{alg:bfs}
\end{algorithm}

\begin{thm}[Correctness of the evolving graph BFS]
Algorithm~\ref{alg:bfs} discovers every active node that is reachable from the
root $(v_1, t_1)$, and $reached[(v, t)]$ is the distance from $(v_1, t_1)$ to
$(v, t)$.
\end{thm}

\proof
Define a static graph $G=(V,E)$ with node set $V = $V^{[1]}$ $\{t\}, t\in\{t_1, t_2, \dots, t_n\}$  such that We could map an evolving graph $G_n$ to a static graph $G$ in such a way that
if there is an edge from $(v, t)$ to $(u,t)$ in $G_n$, there is an edge from $v_t$ to $u_t$ in $G$.
For any two nodes $v_s$ and $v_t$ in $G$, if $s < t$, we add an edge from $v_s$ to $v_t$.
By construction, an active node $(v,t)$ in $G_n$ is in one-to-one correspondence with
a node $v_t$ in $G$.
The forward-neighbors of an active node $(v, t)$ in $G_n$
are in one-to-one correspondence with the forward-neighbors of a node $v_t$ in $G$.
Hence the correctness of the evolving graph BFS follows from that of a static graph.
\endproof

As presented, the BFS over evolving graphs makes no assumptions about how the% adjacency matrices
Suppose $G_n$ is represented by adjacency lists.
Then the number of adjacency lists is the same as the number
of active nodes in $G_n$. So for large $n$, this representation
can use a large amount of memory.
For the graph $G^{[i]} = (V^{[i]}, E^{[i]})$ at time stamp $i$,
we denote the active node set of $V^{[i]}$ as $\tilde V^{[i]}$,
where we have $\tilde V^{[i]} \subseteq V^{[i]}$.
Step $10$ of  Algorithm \ref{alg:bfs} ensures each active node-time stamp pair in $G_n$ is enqueued in $next$ (and then $frontier$) at most once.
And so \texttt{forward-neighbors} loops through each
active node-time stamp pair at most once.
The operations of enqueuing take $O(1)$ time.
Thus the total operations of enqueuing are
$O(\sum_{i=1}^n|\tilde V^{[i]}|)$, where $|\cdot|$ represents the number of elements of a set.  Let $\sum_i |\tilde V^{[i]}| = |\tilde V|$ and $\sum_i |E^{[i]}| = |\hat E|$.
Then the time of scanning
\texttt{forward-neighbors} is
\[
\sum_{i=1}^n\sum_{v \in \tilde V^{[i]}} \texttt{forward-neighbors}(v) =
O(|\hat E|),
\]
by the definition of forward-neighbors.
Hence the total running time is $O(|\hat E|+ |\tilde V|)$.

\section{Formulating the Evolving Graph BFS with Linear Algebra}
\label{sec:representation}
As before, let $G_n = \langle G^{[1]}, G^{[2]}, \ldots, G^{[n]}\rangle$ be an evolving graph with $n$ time stamps. For each graph $G^{[t]}$, we use $A^{[t]}$ to denote
the corresponding adjacency matrix.
\[
A^{[t]}(i,j) =
\begin{cases}
 1 & \mbox{if $(i,j)$ is an edge of $G^{[t]}$,} \\
 0 & \mbox{otherwise.}
\end{cases}
\]
 Each $A^{[t]}$ can be represented by
a $|V|\times |V|$ sparse matrix,
where $|V|$ is the total number of nodes in $G_n$.
We use a sequence of adjacency matrices $A_n = \langle A^{[1]}, A^{[2]}, \ldots, A^{n]}\rangle$ to represent $G_n$.
For example, the evolving graph in Figure \ref{fig:eg_shortest_path}
can be represented as
\[
\left\langle
  \begin{bmatrix}
    0 & 1 & 0 \\
    0 & 0 & 0 \\
    0 & 0 & 0
  \end{bmatrix},~
 \begin{bmatrix}
   0 & 0 & 1 \\
   0 & 0 & 0 \\
   0 & 0 & 0
 \end{bmatrix},~
 \begin{bmatrix}
  0 & 0 & 0 \\
  0 & 0 & 1 \\
  0 & 0 & 0
 \end{bmatrix}
\right\rangle
\]

@@

@@

@@

% \subsection{Search Tree}
% \label{sec:short-temp-path}
%
% We say two nodes $(v_1,t_1)$ and $(v2, t_2)$ are \emph{temporally connected} if
% $(v_1, t_1)$ and $(v2,t_2)$ are connected by a temporal path.
% Notice this notion is symmetric, reflexive but not transitive. For example, suppose
% $t_1 < t_2 < t_3$ and $(v_1, t_1)$ and $(v_3, t_3)$ are temporally connected
% by a temporal path from $(v_1,t_1)$ to $(v_3, t_3)$. Also
% $(v_3, t_3)$ and $(v_2, t_2)$ are temporally connected by a temporal path from
% $(v_2,t_2)$ to $(v_3, t_3)$. But this does not imply $(v_1, t_1)$ and $(v_2, t_2)$ are temporal connected.
%
% Given a root node $(s, t)$, during the execution the evolving-graph BFS builds a search tree, denoted by $T(s,t)$,
% where nodes in $T(s,t)$ are temporally connected.
% \begin{lemma}
% For a node active at different time stamps, say $(s, t_1)$ and $(s,t_2)$ such that $t_1 \leq t_2$,  we have $T(s, t_2) \subseteq T(s, t_1)$, where $T(s,t_1) = T(s,t_2)$ if and only if
% $t_1 = t_2$.
% \end{lemma}
% \proof
% Trivially, $t_1 =t_2$ if and only if $(s,t_1) = (s,t_2)$ if and only if
% $T(s,t_1) = T(s,t_2)$. Now suppose $t_1 < t_2$.
% Let $(v, t)$ be an arbitrary active node that belongs to $T(s, t_2)$. Then there exists a temporal path
% (constructed by the evolving graph BFS) from $(s,t_2)$ to $(v,t)$. Since both
% $(s,t_1)$ and $(s,t_2)$ are active, $(s,t_2)$ is a forward-neighbor of $(s,t_1)$ by definition, i.e., there exists an edge from $(s,t_1)$ to $(s,t_2)$.  Hence there exists a temporal path
% from $(s,t_1)$ to $(v,t)$ and so $(v,t) \in T(s,t_1)$.
% \endproof
